Insight Into Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-09-03 draft.tex
minor stuff, non-image data, sources

\section{\uppercase{Introduction}}
\label{sec:Introduction}
%%Epidemiology is the study of causation of diseases.
Epidemiology is the study of dissemination, causes and results of health-related states and events.
%%
For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) [??] gather as much information as possible about participants to be assessed towards different diseases.
%%
These information are used to determine risk factors for diseases, help people to improve their lifestyle and their health or to support diagnosis of widespread diseases.
%%
Epidemiological research is strongly hypothesis-driven.
%%
More precisely, observations made by clinicians are translated into hypotheses, which are then statistically evaluated using data variables from epidemiological studies.
%%
As a result, the hypothesis will be accepted or rejected.
%%

%For epidemiological research, cohort studies are acquired.
For epidemiological research, cohort studies are carried out.
%%
Since recently, these studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes, such as gender, age and weight.
%%
Back pain is one of the most frequent diseases in the western civilization. 
%%
Epidemiological analysis of back pain aims to characterize the disease by determining risk factors.
%Many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%Since back pain is one of the most frequent diseases in the western civilization, many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%%
Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for the presence of back pain, no automatic classification approach to characterize back pain based on lumbar back spine attributes has been established.
%%
Furthermore, previous analyses mainly focused on a single parameter or a small subset of parameters [Quelle].
%%

In this paper, we present a thorough analysis of the image-derived data from lumbar back spine datasets and analyze their discriminating power for different characteristics of the disease.
%%
For this purpose, we combine data mining algorithms with data visualization techniques.
%%
%Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for possible back pain risk factors.
Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for being back pain risk factors.
%%
\com{Da in der Auflistung oefter vom IVA-Framework die Rede ist: Wollen wir hierzu noch was schreiben, z.B. dass das im VIS-Paper definiert wurde und was wir darunter verstehen; sowie dass diese Arbeit als Erweiterung dieses Frameworks gedacht ist?}

Our contributions are:
\begin{itemize}
\item An IVA-framework for back pain analysis based on image-derived variables of 2,240 subjects \com{Ist diskutabel, so richtig interaktiv ist es nicht. IVA wurde auch noch nicht vorher definiert.}
%%
\item The identification and evaluation of lumbar spine shape properties for back pain diagnosis
%%
\item Detection of correlations between image-based and socio-demographic as well as medical parameters for an improved hypothesis generation
%%
\item Identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework \com{Wieder: IVA wurde noch nicht definiert}
%%
\end{itemize}
%% ---------------------------------------------
\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%%
\noindent Epidemiological reasoning relies on a strict statistically driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item To assess a hypothesis, epidemiologists compile a list of variables depicting it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age.
%%
When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as a \emph{confounder} for prostate cancer.
%%
%These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
\subsection{Epidemiological Data}
Epidemiological data originates from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies, analyzing one specific disease and its influences on the human body.
%%
We focus on data from large scale cohort studies.
%%
These studies aim to collect as much data as possible for each subject.
%%
As a result, these data can be analyzed regarding many diseases and conditions.
%%

@@

Closest to ours is the work of Gla{\ss}er \cite{Glasser2013} and Niemann et al. \cite{Niemann2014}.

@@

Pinhero et al. applied association rule mining to create high-confidence decision rules based on variables such as age, gender and geographic location about the mortality rate in liver cancer data \cite{Pinheiro}.
%%
Decision rules are easily readable and are well suited for classifying medical data. \com{Das frueher erwaehnen}
%%
The results can be easily communicated to domain experts with little knowledge about the algorithm behind the results.
%%
Once these rules, get too complex, analyzing these decision rules becomes complex.
%%
Sekhavat and Hoeber propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map \cite{Sekhavat}.
%%
The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial value and an target value.
%%
The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.
%%
\paragraph{SHIP Data Analysis.}
Klemm et al. analyzed the lumbar spine variability of of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
%%
They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier cluster.
%%
The extracted lumbar spine shape was correlated with subject size.
%%
The data extraction process is based on the detection algorithm and lumbar spine canal extraction presented in this paper.
%%
Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures.
%%
%Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task \cite{Klemm2014BVM}.
%%
\\\\
Unique for our work is the combination of an IVA-approach by observing interesting feature relations in context of image-derived features using multiple decision trees for one visualization.
%%
We abstract the decision tree results similar to \cite{Turkay}, making them comparable in an overview visualization.
%%
\section{The Lumbar Spine Data Set}
\label{sec:MaterialsAndMethod}
%%
\noindent Our approach allows to analyze many variables simultaneously.
%%
Therefore, epidemiologists compiled the data set with a with a wide range of variables possibly correlating with lumbar back pain.
%%
The data set comprises of 6.753 subjects from two cohorts (4.420 from \texttt{SHIP-Trend-0} and 2.333 from \texttt{SHIP-2}).
%%
%It is an active research area to determine these correlations.
%%

%%
\subsection{Non-Image Data}
%\com{Analysis Using Heterogeneous Correlation}
The variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors as sporting activity or nutrition. \com{Ein paar Details hierzu noch - Datentyp, missing data, aufeinander aufbauende parameter}

@@

Figure~\ref{fig:image-parameter-range} shows the distribution of all image-parameters \add{as generalized pairs plot}.
\rem{We created a generalized pairs plot for subject groups derived by dividing them into quantile groups of body-size.}

@@

\subsubsection{Assessing the Information Gain Using the PCA}
%%
To determine the information gain per image-derived variables, we calculated a PCA and compared the loadings per dimension.
%%
The first three principal components explain 75\% of the variance in the image-derived variables.
%%
The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{curvature sagittal}, \emph{curvature angle} and \emph{curvature angle sagittal}.
%%
The second component, adding 16\% of the variance, weigths  \emph{curvature coronal} and \emph{curvature angle coronal}.
%%
The third component explains 12\% of the variance and weights \emph{torsion} and \emph{curvature transverse}.
%%
This supports our prior conclusion about the low information gain of the transverse planes.
%%
\emph{Torsion} also adds little variance to the information space.
%% ------------------------------------------
\subsubsection{Heterogenous Correlations}
%%
We then expanded our focus on correlations of image-derived parameters with all other non-image variables.
%%
We applied a heterogenous correlation technique to derive correlations between all variables is the data set.
%%
The method uses the following correlation metrics for the different type combination:
%%
\begin{itemize}
%%
\item \emph{Pearson product-moment} for two continuous variables, 
\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
\item \emph{Polychoric} correlations for two categorical variables.
\end{itemize}
\com{TODO: Abbildung einer Korrelationsmatrix? Waere sehr gross!}
%%
All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - identical}.
%%
Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
%%
We display the resulting \emph{contingency matrix} using a heat map, mapping correlation values to color brightness with white for 0 and dark blue for 1 \cite{Klemm2014VIS}.
%%
We calculated the contingency matrix for all size groups and looked for correlations between image- and non-image derived variables.
%%
The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the parameter.
%%
Only weak correlations could be found for \emph{Mean Curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
%%
One surprising result was the small correlation with \emph{torsion}, which correlated with almost no variables (p values between 0 and 0.05) and \emph{parkinson} (0.24).
%%

@@

\section{\uppercase{Evaluation of Decision Trees}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal curvature and torsion.
%%
We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-parameter explain non-image variables.
%%
Decision trees are a popular classification method in data mining for creating predictive models.
%%
Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
They work with numerical as well as categorical data.
%%
This allows epidemiologists to interpret the results without having deep knowledge about the algorithm creating the tree.
%%
Readability is only granted for small trees, complex structures with many branches are not desirable.
%%
Too many branches also impose overfitting to the data.
\com{Quelle Decision Tree}

>>>>

\section{\uppercase{Introduction}}
\label{sec:Introduction}
%%Epidemiology is the study of causation of diseases.
Epidemiology is the study of dissemination, causes and results of health-related states and events.
%%
For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) \cite{SHIP} gather as much information as possible about participants to be assessed towards different diseases.
%%
These information are used to determine risk factors for diseases, help people to improve their lifestyle and their health or to support diagnosis of widespread diseases.
%%
Epidemiological research is strongly hypothesis-driven.
%%
More precisely, observations made by clinicians are translated into hypotheses, which are then statistically evaluated using data variables from epidemiological studies.
%%
As a result, the hypothesis will be accepted or rejected.
%%

%For epidemiological research, cohort studies are acquired.
For epidemiological research, cohort studies are carried out.
%%
Since recently, these studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes, such as \emph{gender}, \emph{age} and \emph{weight}.
%%
Back pain is one of the most frequent diseases in the western civilization and gains importance due to the aging society.
%%
Epidemiological analysis of back pain aims to characterize the disease by determining risk factors.
%Many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%Since back pain is one of the most frequent diseases in the western civilization, many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%%
Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for the presence of back pain, no automatic classification approach to characterize back pain based on lumbar back spine attributes has been established.
%%
Furthermore, previous analyses mainly focused on a single parameter or a small subset of parameters.
%%

In this paper, we present a thorough analysis of the image-derived data from lumbar back spine datasets and analyze their discriminating power for different characteristics of the disease.
%%
For this purpose, we combine data mining algorithms with data visualization techniques.
%%
%Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for possible back pain risk factors.
Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for being back pain risk factors.
%%
The workflow is part of an interactive visual analysis (IVA) framework, constructed to enhance the epidemiological workflow with iterative data driven techniques to derive insight \cite{Klemm2014VIS}.
%%
Our contributions are:
\begin{itemize}
\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects
%%
\item The identification and evaluation of lumbar spine shape properties for back pain diagnosis
%%
\item Detection of correlations between image-based and socio-demographic as well as medical parameters for an improved hypothesis generation
%%
\item Identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
%%
\end{itemize}
%% ---------------------------------------------
\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%%
\noindent Epidemiological reasoning relies on a strict statistically driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item To assess a hypothesis, epidemiologists compile a list of variables depicting it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age.
%%
When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as a \emph{confounder} for prostate cancer.
%%
%These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
\subsection{Epidemiological Data}
Epidemiological data originates from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies, analyzing one specific disease and its influences on the human body.
%%
We focus on data from large scale cohort studies.
%%
These studies aim to collect as much data as possible for each subject.
%%
As a result, these data can be analyzed regarding many diseases and conditions.
%%

@@

Closest to ours is the work of Gla{\ss}er \cite{Glasser2013} and Niemann et al. \cite{Niemann2014}, who both use decision rules for their analysis.
Decision rules are easily readable and are well suited for classifying medical data.
The results can be easily communicated to domain experts with little knowledge about the algorithm behind the results.
The effort of analyzing the decision rules increases with decision rule complexity.

@@

Pinhero et al. applied association rule mining to create high-confidence decision rules based on variables such as age, gender and geographic location about the mortality rate in liver cancer data \cite{Pinheiro}.
%%
Sekhavat and Hoeber propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map \cite{Sekhavat}.
%%
The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial value and an target value.
%%
The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.
%%
\paragraph{SHIP Data Analysis.}
Klemm et al. analyzed the lumbar spine variability of of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
%%
They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier cluster.
%%
The extracted lumbar spine shape was correlated with subject size.
%%
The data extraction process is based on the detection algorithm and lumbar spine canal extraction presented in this paper.
%%
Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures.
%%
%Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task \cite{Klemm2014BVM}.
%%
\\\\
Unique for our work is the combination of an IVA-approach by observing interesting feature relations in context of image-derived features using multiple decision trees for one visualization.
%%
We abstract the decision tree results similar to \cite{Turkay}, making them comparable in an overview visualization.
%%
\section{The Lumbar Spine Data Set}
\label{sec:MaterialsAndMethod}
%%
\noindent Our approach allows to analyze many variables simultaneously.
%%
Therefore, epidemiologists compiled the data set with a with a wide range of variables possibly correlating with lumbar back pain.
%%
The data set comprises of 6.753 subjects from two cohorts (4.420 from \texttt{SHIP-Trend-0} and 2.333 from \texttt{SHIP-2}).
%%
%It is an active research area to determine these correlations.
%%

%%
\subsection{Non-Image Data}
%\com{Analysis Using Heterogeneous Correlation}
The variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors as sporting activity or nutrition.
%%
Without the 9 image-derived parameter, the data set comprises of 134 variables:
\begin{itemize}
	\item 21 metric variables, mostly describing somatometric variables and markers retrieved from blood analysis
	\item 113 categorical variables divided into
	\begin{itemize}
		\item 43 dichotomous variables, mostly indicating the presence of a disease
		\item 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
	\end{itemize}
\end{itemize}
%%
Many variables comprise missing data.
%%
Some variables are follow-up questions, covering the treatment of a disease.
%%
Others are exclusive to subgroups, such as women-specific variables, e.g. menstrual status.
%%
Sparse variables are statistically less resilient and need to be treated with care in the analysis.

@@

Figure~\ref{fig:image-parameter-range} shows the distribution of all image-parameters as generalized pairs plot.

@@

\subsubsection{Assessing the Information Gain Using the PCA}
%%
To determine the information gain per image-derived variables, we calculated a PCA and compared the loadings per dimension.
%%
The first three principal components explain 75\% of the variance in the image-derived variables.
%%
The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{curvature sagittal}, \emph{curvature angle} and \emph{curvature angle sagittal}.
%%
The second component, adding 16\% of the variance, weigths  \emph{curvature coronal} and \emph{curvature angle coronal}.
%%
The third component explains 12\% of the variance and weights \emph{torsion} and \emph{curvature transverse}.
%%
This supports our prior conclusion about the low information gain of the transverse planes.
%%
\emph{Torsion} also adds little variance to the information space.
%% ------------------------------------------
\subsubsection{Heterogenous Correlations}
%%
We then expanded our focus on correlations of image-derived parameters with all other non-image variables.
%%
We applied a heterogenous correlation technique to derive correlations between all variables is the data set.
%%
The method uses the following correlation metrics for the different type combination:
%%
\begin{itemize}
%%
\item \emph{Pearson product-moment} for two continuous variables, 
\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
\item \emph{Polychoric} correlations for two categorical variables.
\end{itemize}
%%
All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - identical}.
%%
Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
%%
We display the resulting \emph{contingency matrix} using a heat map, mapping correlation values to color brightness with white for 0 and dark blue for 1 \cite{Klemm2014VIS}.
%%
We calculated the contingency matrix for all size groups and looked for correlations between image- and non-image derived variables.
%%
The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the parameter.
%%
Only weak correlations could be found for \emph{Mean Curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
%%
One surprising result was the small correlation with \emph{torsion}, which correlated with almost no variables (p values between 0 and 0.05) and \emph{parkinson} (0.24).
%%

@@

\section{\uppercase{Evaluation of Decision Trees}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal curvature and torsion.
%%
We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-parameter explain non-image variables.
%%
Decision trees are a popular classification method in data mining for creating predictive models.
%%
Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
They work with numerical as well as categorical data.
%%
This allows epidemiologists to interpret the results without having deep knowledge about the algorithm creating the tree.
%%
Readability is only granted for small trees, complex structures with many branches are not desirable.
%%
Too many branches also impose overfitting to the data \cite{DecisionTree}.
